Space for WHAM: a multi-region, multi-stock generalization of the Woods Hole Assessment Model with an application to black sea bass

### 2024-11-19 paper/paper.tex
saving work

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={A multi-stock, multi-region state-space age-structured assessment model with an application to black sea bass},
  pdfauthor={Timothy J. Miller1; Kierstin Curti; Alex Hansell},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage{url}
\usepackage{setspace}
%\singlespacing
%\onehalfspacing
\doublespacing
\usepackage{lineno}
\linenumbers
\usepackage[belowskip=0pt,aboveskip=0pt]{caption}
\usepackage{relsize}
\usepackage{float}
% \usepackage[section]{placeins}
\usepackage{lscape}
\usepackage{longtable}
\usepackage{amsmath,rotating}
\usepackage[scanall]{psfrag}
\usepackage{bm}
\usepackage{caption,graphics}
\usepackage{graphicx}
%\usepackage{natbib}
%\usepackage[nottoc]{tocbibind}
%\usepackage{indentfirst}
\usepackage{sectsty}
\usepackage{color}
\usepackage{fancyhdr}
\usepackage{xspace}
\usepackage{textcomp}
\usepackage{upgreek}

@@

\title{A multi-stock, multi-region state-space age-structured assessment
model with an application to black sea bass}
\author{Timothy J. Miller\textsuperscript{1} \and Kierstin
Curti \and Alex Hansell}
\date{16 September, 2024}

@@

Why is this model needed. Models ignoring movement can provide biased
catch advice. Using random effects for time-varying processes is
considered best practice.

The Woods Hole Assessment Model (WHAM) is an R package developed at the
Northeast Fisheries Science Center
(\url{https://timjmiller.github.io/wham}, Miller and Stock 2020; Stock
and Miller 2021). Primary inputs and observation types are similar to
ASAP and there is functionality built into WHAM to move information from
ASAP input files to R. WHAM can be configured to fit SCAA models often
identically to ASAP, so that existing assessments in the U.S. Northeast
can be replicated and tested against state-space models with process
errors and environmental effects in a single framework. WHAM models are
built using the Template Model Builder package (TMB, Kristensen et al.
2016) which provides efficient fitting of models that include random
effects similar to the State-space Assessment Model (SAM, Nielsen and
Berg 2014), which is used to manage many stocks within ICES. WHAM is
currently used to manage 9 stocks in the NEUS.

The Woods Hole Assessment Model (WHAM) software package is maintained
and developed at the Northeast Fisheries Science Center to enable
state-space stock assessments, i.e.~where processes such as the annual
transitions in numbers-at-age (survival), natural mortality,
selectivity, and catchability are treated as time- and(or) age-varying
random effects. WHAM can also be configured without random effects as a
traditional statistical catch-at-age model in order to bridge from
current assessments which use Age-Structured Assessment Program (ASAP).
Here, we present an extension of WHAM (Multi-WHAM) to model multiple
stocks and(or) regions and allow movement of stocks among regions.

It has been reviewed in the literature and simulation tested (Stock et
al. 2021; Stock and Miller 2021). It has also been used as the operating
model in the index-based assessment methods research track assessment
(Legault et al. 2023) and is also the main tool for simulation studies
in the research track on state-space assessment models.

A review of the essential features for next-generation stock assessments
concluded that only WHAM and a similar State-Space Assessment Method
(Nielsen and Berg 2014) model random effects correctly (Punt et
al.~2020). WHAM is now being used as the management model for 9 stocks
in the northeast U.S., but the standard version of WHAM can only be
applied to a single stock and area (Miller and Stock 2020).

Here, we describe a multi-stock and multi-region extension of the WHAM
package (version 2.0 \url{https://github.com/timjmiller/wham/tree/lab}).
Numerous new configuration options are also included in this extension
that can be useful even for single-stock or single region models. Any
number of stocks or regions can be configured in this version of WHAM.
The multi-stock extension was developed to accommodate movement of
stocks or stock components in anticipation of black sea bass and
Atlantic cod stocks where new modeling frameworks where recently
examined through Northeast research track assessment process (cite RT
docs). We demonstrate capabilities of the new version of WHAM with black
sea bass and evaluate evidence for alternative hypotheses of temporal
variation in natural mortality and effects of bottom temperature on
recruitment and natural mortality of age 1 individuals.

@@

\hypertarget{model-description}{%
\subsection*{Model description}\label{model-description}}
\addcontentsline{toc}{subsection}{Model description}

Many of the options and equations of version 2.0 are are the same as
those described by Stock and Miller (2021), so we will only describe
extensions and differences that have occurred since the first
description of WHAM. The new version of WHAM can model multiple stocks,
but survival, movement, harvest and natural mortality is are tracked for
each stock. Therefore, much of the description below assumes a specific
stock \(s\), but this subscript is only explicitly used when necessary.

@@

Because individuals may be alive in one of several regions or harvested
in one of several fleets, it is helpful to consider these as distinct
categories or states and treat the number of individuals occurring in
each category over time as a multi-state model (reference). Multi-state
models use a probability transition matrix (PTM) that describes the
probability of individuals living and moving among regions or dying due
to fishing or natural mortality over a time interval \(i\) with duration
\(\delta_i\) in year \(y\) for individuals at age \(a\) on January 1.
Each row and column of the PTM correspond to one the states: alive in
region \(r\), dead in fleet \(f\), or dead from natural causes. The
probabilities in each row sum to unity and assume an individual is in
the corresponding state at the beginning of the interval. Given \(n_R\)
regions and \(n_F\) fleets, the square PTM (\(n_R + n_F + 1\) rows and
columns) as a function of sub-matrices is \begin{equation}\label{eq:ptm}
  \mathbf{P}_{y,a,i} = \begin{bmatrix}
    \mathbf{O}_{y,a,i} & \mathbf{H}_{y,a,i} & \mathbf{D}_{y,a,i} \\
    0 & \mathbf{I}_{H} & 0\\
    0 & 0 & 1
  \end{bmatrix}
\end{equation} where \begin{equation*}
  \mathbf{O}_{y,a,i} = 
  \begin{bmatrix}
    O_{y,a,i}(1,1) & \cdots & O_{y,a,i}(1,n_R) \\
    \vdots & \ddots & \vdots \\
    O_{y,a,i}(n_R,1) & \cdots & O_{y,a,i}(n_R,n_R)
  \end{bmatrix}
\end{equation*} is the \(n_R \times n_R\) matrix defining survival and
occurring in each region at the end of the interval, \begin{equation*} 
  \mathbf{H}_{y,a,i} = 
  \begin{bmatrix}
    H_{y,a,i}(1,1) & \cdots & H_{y,a,i}(1,n_F) \\
    \vdots & \ddots & \vdots \\
    H_{y,a,i}(n_R,1) & \cdots & H_{y,a,i}(n_R,n_F)
  \end{bmatrix}
\end{equation*} is the \(n_R \times n_F\) matrix defining probabilities
of being captured in each fleet during the interval, and
\(\mathbf{D}_{y,a,i}\) is the \(n_R x 1\) matrix of probabilities of
dying due to natural mortality during the interval. We have the identity
matrix \(\mathbf{I}_{H}\) for the states for capture by each fleet and a
1 for the state for natural mortality because the probabilities of being
in one of the mortality states given starting the interval in that state
is unity.

WHAM uses these PTMs to model abundance proportions in each state rather
than true probabilities where numbers in each state would be multinomial
distributed. The PTMs determine the expected numbers 1) in each state on
January 1 of year \(t+1\) at age \(a+1\) given the abundances at age
\(a\) on January 1 of year \(t\), 2) captured over the year in each
fleet, 3) available to each index, and 4) alive at the time and in the
region where spawning occurs.

@@

As in previous versions of WHAM, the transitions in numbers at age from
one year to another after recruitment can be treated deterministically
or as functions of random effects. The predicted numbers at age in year
\(y\) at age \(a\) for a given stock are vector analogs
(\(\mathbf{N}_{O,y,a}\)) of the equations for numbers at age in the
standard WHAM model (Stock and Miller 2021). For ages
\(a = 2,\ldots, A-1\), where \(A\) is the plus group, the expected
number alive in each region at the beginning of the following year and
next age class age can be obtained from the first \(n_R\) elements of
the vector \[\mathbf{P}_{y-1,a-1}' \mathbf{N}_{y-1,a-1}.\] The numbers
alive in each region can also be modeled more simply using the
sub-matrix \(\mathbf{O}_{y,a}\). The general model for the transitions
in abundance at age is \begin{equation*}
\log\left(\mathbf{N}_{O,y,a}\right)|\mathbf{N}_{O,y-1,a-1} =  \log\left(\mathbf{O}_{y-1,a-1}' \mathbf{N}_{O,y-1,a-1}\right) + \boldsymbol{\varepsilon}_{y,a}
\end{equation*} for ages \(a = 2,\ldots, A-1\), and for the plus group
\begin{equation*}
\log\left(\mathbf{N}_{O,y,A}\right)|\mathbf{N}_{O,y-1,a-1},\mathbf{N}_{O,y-1,A} = \log\left(\mathbf{O}_{y-1,A-1}' \mathbf{N}_{O,y-1,A-1} + \mathbf{O}_{y-1,A}' \mathbf{N}_{O,y-1,A}\right) + \boldsymbol{\varepsilon}_{y,A}.
\end{equation*} When the transitions in abundance at age are treated
deterministically, \(\boldsymbol{\varepsilon}_{y,a} = 0\). The stock-
and region-specific errors \(\boldsymbol{\varepsilon}_{y,a}\) are
independent, but the same correlation structures as previous versions
are possible across ages and years for a given stock and region. When
there is autocorrelation with age, WHAM now assumes this applies only to
ages \(a>1\) by default so that recruitment random effects are
independent of those for for the annual transitions of older age
classes. So the general covariance structure for a given stock at ages
\(a>1\) in region \(r\) is \begin{equation*}
  Cov\left(\epsilon_{y,a,r},\epsilon_{y',a',r}\right) =   \frac{\rho_{N,\text{age},r}^{|a-a'|}\rho_{N,\text{year},r}^{|y-y'|}\sigma^2_{N,r}}{\left(1 -  \rho_{N,\text{age},r}^2\right)\left(1 - \rho_{N,\text{year},r}^2\right)}.
\end{equation*}

@@

The initial abundances at age can also be treated as independent or
autocorrelated (with age) random effects. Defining the vector of initial
abundance at age in region \(r\) as \(\mathbf{N}_{O,1,r}\), the general
model is
\[\log \mathbf{N}_{O,1,r} = \theta_{N_1,r} + \boldsymbol{\varepsilon}_{N_1,r}\]
where
\[  Cov\left(\varepsilon_{N_1,a,r},\varepsilon_{N_1,a',r}\right) = \frac{\rho_{N_1,r}^{|a-a'|}\sigma^2_{N_1,r}} {\left(1-\rho_{N_1,r}^2\right)}.\]

@@

For each season, there are at most \(n_R-1\) parameters determining
movement among regions given starting an the season in region \(r\) in
either the sequential or simultaneous configurations. Movement
parameters are estimated on a transformed scale via a link function
\(g(\cdot)\). If survival and movement are occur simultaneously, the
parameters are estimated a log link function is used and if they are
separable, an additive logit link function (like a multinomial
regression) is used. On the transformed scale, the general model for the
movement parameter from region \(r\) to \(r'\) in season \(i\) and year
\(y\) for individuals of age \(a\) is a linear function of both random
and environmental effects: \begin{equation*}
  g(\mu_{r\rightarrow r',y,a,i}) = \theta_{r\rightarrow r',i} + \epsilon_{r\rightarrow r',y,a,i} + \sum^{n_E}_{k=1} \beta_{r \rightarrow r',a,i,k} E_{k,y}.
\end{equation*} The random effects \(\epsilon_{r\rightarrow r',y,a,i}\)
are season-, and(or) region-to-region-specific and modeled most
generally as a two-dimensional first-order autoregressive random effects
with age and(or) year where the covariance is \begin{equation*}
  Cov\left(\epsilon_{r\rightarrow r',y,a,i},\epsilon_{r\rightarrow r',y',a',i}\right) =   \frac{\rho_{r\rightarrow r',\text{age},i}^{|a-a'|}\rho_{r\rightarrow r',\text{year},i}^{|y-y'|}\sigma^2_{r\rightarrow r',i}}{\left(1 -  \rho_{r\rightarrow r',\text{age},i}^2\right)\left(1 - \rho_{r\rightarrow r',\text{year},i}^2\right)}
\end{equation*} similar to how WHAM models variation in survival,
natural mortality, and selectivity. Effects of covariate \(E_k\) can be
age-, season-, and(or) region-to-region-specific
\(\beta_{r\rightarrow r',a,i,k}\) and the same orthogonal polynomial
options in the previous versions of WHAM for effects on recruitment and
natural mortality are available.

@@

Natural mortality options have been expanded in WHAM. When not
estimated, (mean) mortality rates may be stock-, region-, and
age-specific. When random effects are used, the same autocorrelation
structures with age and year as described by Stock and Miller (2021) can
be configured for a given stock and region. Any environmental covariate
effects can be stock-, region-, and age-specific. So the general model
for natural mortality is \begin{equation*}
  \log M_{y,a,r} = \theta_{M,r} + \epsilon_{M,r,y,a} + \sum^{n_E}_{k=1} \beta_{M,r,a,k} E_{k,y}.
\end{equation*} The general covariance structure for random effects are
modeled most generally as a two-dimensional first-order autoregressive
random effects with age and(or) year where the covariance is
\begin{equation*}
  Cov\left(\epsilon_{M,y,a,r},\epsilon_{M,y',a',r}\right) =   \frac{\rho_{M,\text{age},r}^{|a-a'|}\rho_{M,\text{year},r}^{|y-y'|}\sigma^2_{M,r}}{\left(1 -  \rho_{M,\text{age},r}^2\right)\left(1 - \rho_{M,\text{year},r}^2\right)}.
\end{equation*}

@@

The first year being modeled for the population is 1989 and the fishery
and index data used in the model span from 1989 to 2021. There are two
stock components (north and south) modeled as separate populations that
spawn and recruit in respective regions separated by the Hudson Canyon.
We have observations for each of four total fishing fleets, where two
fishing fleets (Recreational and Commercial) operate in each region.

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Space for WHAM: a multi-region, multi-stock generalization of the Woods Hole Assessment Model with an application to black sea bass},
  pdfauthor={Timothy J. Miller1; Kierstin Curti; Alex Hansell},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage{url}
\usepackage{setspace}
%\singlespacing
%\onehalfspacing
\doublespacing
\usepackage{lineno}
\linenumbers
\usepackage[belowskip=0pt,aboveskip=0pt]{caption}
\usepackage{relsize}
\usepackage{float}
% \usepackage[section]{placeins}
\usepackage{lscape}
\usepackage{longtable}
\usepackage{amsmath,rotating}
\usepackage[scanall]{psfrag}
\usepackage{bm}
\usepackage{caption,graphics}
\usepackage{graphicx}
%\usepackage{natbib}
%\usepackage[nottoc]{tocbibind}
%\usepackage{indentfirst}
\usepackage{sectsty}
\usepackage{color}
\usepackage{fancyhdr}
\usepackage{xspace}
\usepackage{textcomp}
\usepackage{upgreek}

@@

\title{Space for WHAM: a multi-region, multi-stock generalization of the
Woods Hole Assessment Model with an application to black sea bass}
\author{Timothy J. Miller\textsuperscript{1} \and Kierstin
Curti \and Alex Hansell}
\date{14 November, 2024}

@@

A state-space statistical approach and maximum marginal likelihood or
Bayesian fitting of stock assessment models allows estimation of time-
and age-varying population attributes as random effects Miller et al.
(2016b). This estimation approach is considered an essential feature of
gold-standard assessment models that we use in tactical management of
commercially important fish stocks (Punt et al. 2020). The State-space
Assessment Model (SAM, Nielsen and Berg 2014) continues to be developed
and remains widely used within ICES to assess European fish stocks.
Various state-space models are being used to manage cod and plaice
stocks in the waters of Eastern Canada (Perreault et al. 2020; Varkey et
al. 2022) and to the south, the Woods Hole Assessment Model (WHAM, Stock
and Miller 2021) is now used to assess many fish stocks in the Northwest
Atlantic Ocean (NEFSC 2022a, 2022b; NEFSC 2024).

WHAM is an R package developed and maintained at NOAA's Northeast
Fisheries Science Center (\url{https://timjmiller.github.io/wham},
Miller and Stock 2020; Stock and Miller 2021). WHAM can be configured to
fit a wide range of age-structured models from traditional statistical
catch-at-age models without any random effects to models with several
time and age varying process errors and possibly effects of
environmental covariates on various demographic parameters. Like SAM,
WHAM models are built using the Template Model Builder package (TMB,
Kristensen et al. 2016) which provides a computationally efficient means
of fitting an extremely wide class of models with random effects. WHAM
has undergone active development since its creation and includes random
effects options for in recruitment, inter-annual transitions numbers at
age (hereafter referred to as ``survival''), fishery and index
selectivity, natural mortality, and catchability.

However, WHAM has up to now only allowed models with one stock
(component) and a single region and without seasonal changes in stock
dynamics. Using such models for stocks that have subcomponents with
varying seasonal movement can provide incorrect inferences and poor
management advice (Ying et al. 2011; Cao et al. 2014; Bosley et al.
2022). Furthermore, the ability to account for spatial structure and
model multiple stocks are also important features of leading-edge
assessment modeling frameworks (Punt et al. 2020). We describe here the
implementation of these features and other extensions since Stock and
Miller (2021) in WHAM version 2.0. Many of these new configuration
options can be useful whether modeling 1 or more stocks and regions.

This extension of WHAM was developed in concert with a stock assessment
for black sea bass through the NEFSC research track assessment process
where new modeling frameworks were recently examined and modeling
multiple stocks or stock components simultaneously was of interest
(NEFSC 2023). \textcolor{red}{[Text on black sea bass here]} We apply
WHAM 2.0 to two stock components of black sea bass off the coast of the
NEUS and evaluate evidence for alternative hypotheses of temporal
variation and effects specifically of bottom temperature on recruitment
and natural mortality of age 1 individuals.

@@

\hypertarget{wham-description}{%
\subsection*{WHAM description}\label{wham-description}}
\addcontentsline{toc}{subsection}{WHAM description}

Many of the options and equations of WHAM version 2.0 are the same as
those in Stock and Miller (2021), so we will only describe extensions
and differences that have occurred since their first description of
WHAM. The new version of WHAM can model multiple stocks and survival,
movement, harvest and natural mortality are tracked for each stock. Much
of the description below is for a specific stock \(s\), but, for
simplicity, this subscript is implicit except when necessary.

@@

Because individuals may be alive in one of several regions or harvested
in one of several fleets, it is helpful to consider these as distinct
categories or states and treat the number of individuals occurring in
each category over time as a multi-state model. Approaches to modeling
transitions among these tates may treat time discretely (e.g., Arnason
1972; Schwarz et al. 1993) or continuously (e.g., Hearn et al. 1987;
Commenges 1999; Andersen and Keiding 2002). Multi-state models can
define a probability transition matrix (PTM) that describes the
probability of individuals occurring in different states at the end of a
time interval \(\delta\), conditional on being in each of those state at
beginning of the interval. For fish populations these states would be
defined as being alive in a particular region or being dead due to
fishing from a particular fleet or natural mortality. The time interval
\(i\) with duration \(\delta_i\) would be a season and the PTMs would be
uniquely defined for each stock by season \(i\), year \(y\), and age
\(a\) on January 1. Each row and column of the PTM correspond to one the
states: alive in region \(r\), dead in fleet \(f\), or dead from natural
causes. The probabilities in each row sum to unity and assume an
individual is in the corresponding state at the beginning of the
interval. Given \(n_R\) regions and \(n_F\) fleets, the square PTM
(\(n_R + n_F + 1\) rows and columns) as a function of sub-matrices is
\begin{equation}\label{eq:ptm}
  \mathbf{P}_{y,a,i} = \begin{bmatrix}
    \mathbf{O}_{y,a,i} & \mathbf{H}_{y,a,i} & \mathbf{D}_{y,a,i} \\
    0 & \mathbf{I}_{H} & 0\\
    0 & 0 & 1
  \end{bmatrix}
\end{equation} where \begin{equation*}
  \mathbf{O}_{y,a,i} = 
  \begin{bmatrix}
    O_{y,a,i}(1,1) & \cdots & O_{y,a,i}(1,n_R) \\
    \vdots & \ddots & \vdots \\
    O_{y,a,i}(n_R,1) & \cdots & O_{y,a,i}(n_R,n_R)
  \end{bmatrix}
\end{equation*} is the \(n_R \times n_R\) matrix defining survival and
occurring in each region at the end of the interval, \begin{equation*} 
  \mathbf{H}_{y,a,i} = 
  \begin{bmatrix}
    H_{y,a,i}(1,1) & \cdots & H_{y,a,i}(1,n_F) \\
    \vdots & \ddots & \vdots \\
    H_{y,a,i}(n_R,1) & \cdots & H_{y,a,i}(n_R,n_F)
  \end{bmatrix}
\end{equation*} is the \(n_R \times n_F\) matrix defining probabilities
of being captured in each fleet during the interval, and
\(\mathbf{D}_{y,a,i}\) is the \(n_R x 1\) matrix of probabilities of
dying due to natural mortality during the interval. We have the identity
matrix \(\mathbf{I}_{H}\) for the states for capture by each fleet and a
1 for the state for natural mortality because the probabilities of being
in one of the mortality states given starting the interval in that state
is unity (no zombies allowed).

WHAM uses these PTMs to model abundance proportions in each state rather
than true probabilities where numbers in each state would be multinomial
distributed as in a model for tagging data where fates of individual
fish are assumed independent. The PTMs determine the expected numbers 1)
in each state on January 1 of year \(t+1\) at age \(a+1\) given the
abundances at age \(a\) on January 1 of year \(t\), 2) captured over the
year in each fleet, 3) available to each index, and 4) alive at the time
and in the region where spawning occurs.

@@

As in previous versions of WHAM, the transitions in numbers at age from
one year to another after recruitment can be treated deterministically
or as functions of random effects. The predicted numbers at age in year
\(y\) at age \(a\) for a given stock are vector analogs
(\(\mathbf{N}_{O,y,a}\)) of the equations for numbers at age in the
standard WHAM model (Stock and Miller 2021). For ages
\(a = 2,\ldots, A-1\), where \(A\) is the plus group, the expected
number alive in each region at the beginning of the following year and
next age class age can be obtained from the first \(n_R\) elements of
the vector \[\mathbf{P}_{y-1,a-1}' \mathbf{N}_{y-1,a-1}.\] The numbers
alive in each region can also be modeled more simply using the
sub-matrix \(\mathbf{O}_{y,a}\). The general model for the transitions
in abundance at age is \begin{equation*}
\log\left(\mathbf{N}_{O,y,a}\right)|\mathbf{N}_{O,y-1,a-1} =  \log\left(\mathbf{O}_{y-1,a-1}' \mathbf{N}_{O,y-1,a-1}\right) + \boldsymbol{\varepsilon}_{y,a}
\end{equation*} for ages \(a = 2,\ldots, A-1\), and for the plus group
\begin{equation*}
\log\left(\mathbf{N}_{O,y,A}\right)|\mathbf{N}_{O,y-1,a-1},\mathbf{N}_{O,y-1,A} = \log\left(\mathbf{O}_{y-1,A-1}' \mathbf{N}_{O,y-1,A-1} + \mathbf{O}_{y-1,A}' \mathbf{N}_{O,y-1,A}\right) + \boldsymbol{\varepsilon}_{y,A}.
\end{equation*} When the transitions in abundance at age are treated
deterministically, \(\boldsymbol{\varepsilon}_{y,a} = 0\). The stock-
and region-specific errors \(\boldsymbol{\varepsilon}_{y,a}\) are
independent, but the same correlation structures as previous versions
are possible across ages and years for a given stock and region. When
there is autocorrelation with age, WHAM now assumes this applies only to
ages \(a>1\) by default so that recruitment random effects are
independent of those for for the annual transitions of older age
classes. So the general covariance structure for a given stock at ages
\(a>1\) in region \(r\) is that of a two-dimensional first-order
autoregressive (2DAR1) process \begin{equation*}
  Cov\left(\epsilon_{y,a,r},\epsilon_{y',a',r}\right) =   \frac{\rho_{N,\text{age},r}^{|a-a'|}\rho_{N,\text{year},r}^{|y-y'|}\sigma_{N,a,r}\sigma_{N,a',r}}{\left(1 -  \rho_{N,\text{age},r}^2\right)\left(1 - \rho_{N,\text{year},r}^2\right)} 
\end{equation*} and that for age 1 is just AR1 across years
\begin{equation*}
  Cov\left(\epsilon_{y,1},\epsilon_{y',1}\right) =   \frac{\rho_{N,1,\text{year}}^{|y-y'|}\sigma^2_{N,1}}{1 - \rho_{N,1,\text{year}}^2}.
\end{equation*} Since recruitment for a given stock currently only
occurs in one region \(r_s\) there is only a single time-varying
recruitment random effect for each stock.

@@

The initial abundances at age can also be treated as independent or as
AR1 random effects. Defining the vector of initial abundance at age in
region \(r\) as \(\mathbf{N}_{O,1,r}\), the general model is
\[\log \mathbf{N}_{O,1,r} = \theta_{N_1,r} + \boldsymbol{\varepsilon}_{N_1,r}\]
where
\[  Cov\left(\varepsilon_{N_1,a,r},\varepsilon_{N_1,a',r}\right) = \frac{\rho_{N_1,r}^{|a-a'|}\sigma^2_{N_1,r}} {\left(1-\rho_{N_1,r}^2\right)}.\]

@@

For each season, there are at most \(n_R-1\) parameters determining
movement among regions given starting an the season in region \(r\) in
either the sequential or simultaneous configurations. Movement
parameters are estimated on a transformed scale via a link function
\(g(\cdot)\). If survival and movement are occur simultaneously, the
parameters are estimated a log link function is used and if they are
separable, an additive logit link function (like a multinomial
regression) is used. On the transformed scale, the general model for the
movement parameter from region \(r\) to \(r'\) in season \(i\) and year
\(y\) for individuals of age \(a\) is a linear function of both random
and environmental effects: \begin{equation*}
  g(\mu_{r\rightarrow r',y,a,i}) = \theta_{r\rightarrow r',i} + \epsilon_{r\rightarrow r',y,a,i} + \sum^{n_E}_{k=1} \beta_{r \rightarrow r',a,i,k} E_{k,y}.
\end{equation*} The random effects \(\epsilon_{r\rightarrow r',y,a,i}\)
are season-, and(or) region-to-region-specific and modeled most
generally as 2DAR1 random effects with age and(or) year where the
covariance is \begin{equation*}
  Cov\left(\epsilon_{r\rightarrow r',y,a,i},\epsilon_{r\rightarrow r',y',a',i}\right) =   \frac{\rho_{r\rightarrow r',\text{age},i}^{|a-a'|}\rho_{r\rightarrow r',\text{year},i}^{|y-y'|}\sigma^2_{r\rightarrow r',i}}{\left(1 -  \rho_{r\rightarrow r',\text{age},i}^2\right)\left(1 - \rho_{r\rightarrow r',\text{year},i}^2\right)}
\end{equation*} similar to how WHAM models variation in survival,
natural mortality, and selectivity. Effects of covariate \(E_k\) can be
age-, season-, and(or) region-to-region-specific
\(\beta_{r\rightarrow r',a,i,k}\) and the same orthogonal polynomial
options in the previous versions of WHAM for effects on recruitment and
natural mortality are available.

@@

Natural mortality options have been expanded in WHAM. When not
estimated, (mean) mortality rates may be stock-, region-, and
age-specific. When random effects are used, the same 2DAR1 structure
with age and year as described by Stock and Miller (2021) can be
configured for a given stock and region. Any environmental covariate
effects can be stock-, region-, and age-specific. So the general model
for natural mortality is \begin{equation*}
  \log M_{y,a,r} = \theta_{M,r} + \epsilon_{M,r,y,a} + \sum^{n_E}_{k=1} \beta_{M,r,a,k} E_{k,y}.
\end{equation*} The general covariance structure for random effects are
modeled most generally as 2DAR1 random effect with age and(or) year
where the covariance is \begin{equation*}
  Cov\left(\epsilon_{M,y,a,r},\epsilon_{M,y',a',r}\right) =   \frac{\rho_{M,\text{age},r}^{|a-a'|}\rho_{M,\text{year},r}^{|y-y'|}\sigma^2_{M,r}}{\left(1 -  \rho_{M,\text{age},r}^2\right)\left(1 - \rho_{M,\text{year},r}^2\right)}.
\end{equation*}

@@

The first year being modeled for the population is 1989 and the fishery
and index data used in the model span from 1989 to 2021. The north and
south stock components are modeled as separate populations that spawn
and recruit in respective regions. We have observations for each of four
total fishing fleets, where two fishing fleets (Recreational and
Commercial) operate in each region.
